On the entropy of isothermal dusty gas: application to protoplanetary disks

### 2016-08-02 paper.tex
adding fixZ results

\title{On the entropy of 
  isothermal dusty gas: application to protoplanetary disks}  
\author{Min-Kai Lin\altaffilmark{1}}
\affil{Department of Astronomy and Steward Observatory, \\
University of
  Arizona, 933 North Cherry Avenue, Tucson, AZ 85721, USA}
\altaffiltext{1}{Steward Theory Fellow}
\email{minkailin@email.arizona.edu}

\begin{abstract}
  Small dust particles embedded in gaseous protoplanetary disks are subject
  to strong dust-gas friction, causing them to almost follow the gas
  flow. This is similar to the conservation of entropy following an
  adiabatic fluid.  We show that in the strong drag
  limit that an isothermal dusty gas indeed behaves like an  
  adiabatic pure gas, and finite dust-gas coupling may be
  regarded as an effective 
  heating/cooling. We exploit this equivalence 
  to 
\begin{inparaenum}[1)]
\item
  define the entropy of isothermal dusty gas; and 
\item 
  generalize previous, well-known gaseous instabilities in
  protoplanetary disks to include dust dynamics. 
\end{inparaenum}
This includes the axisymmetric stability of strictly isothermal dusty
gas, from which we place quantitative constraints on the sharpness of 
dust edges that can persist in protoplanetary disks. We also extend
the vertical shear instability (VSI) in locally isothermal disks to include
dust. We find that  dust-loading reduces VSI growth rates. The presence of
strongly-coupled dust particles preferentially stabilize VSI modes with small radial
wavelengths, as well as `surface modes' concentrated near the
disk boundaries. We briefly discuss potential applications of the 
equivalence between isothermal dusty gas and standard hydrodynamics
to provide thermodynamical interpretations of  other dusty instabilities in protoplanetary disks. 
\end{abstract}

@@

\input{model}
\input{linear}
\input{result}
%\input{linear_adia}
%\input{numerical}
%\input{application} 
%\input{summary}

>>>>

\title{On the entropy of 
  isothermal dusty gas: application to protoplanetary disks}  
\author{Min-Kai Lin\altaffilmark{1} and Andrew N. Youdin}
\affil{Department of Astronomy and Steward Observatory, \\
University of
  Arizona, 933 North Cherry Avenue, Tucson, AZ 85721, USA}
\altaffiltext{1}{Steward Theory Fellow}
\email{minkailin@email.arizona.edu}

\begin{abstract}
  Small solids embedded in gaseous protoplanetary disks are subject
  to strong dust-gas friction. Tightly-coupled dust particles
  therefore almost follow the gas flow. This is analogous to the
  conservation of entropy 
  following an adiabatic fluid.  We show that in the strong drag
  limit that an isothermal dusty gas indeed behaves like an  
  adiabatic pure gas, and finite dust-gas coupling may be
  regarded as an effective 
  heating/cooling. We exploit this equivalence 
  to 
\begin{inparaenum}[1)]
\item
  define the entropy of isothermal dusty gas; and 
\item 
  generalize previous, well-known gaseous instabilities in
  protoplanetary disks to include dust dynamics. 
\end{inparaenum}
This includes the axisymmetric stability of strictly isothermal dusty
gas, allowing us to constrain the sharpness of 
dust edges that can persist in protoplanetary disks. We also extend
the vertical shear instability (VSI) in locally isothermal disks to include
dust. We find that  dust-loading reduces VSI growth rates, and
preferentially stabilizes VSI modes with small radial 
wavelengths, as well as `surface modes' concentrated near the
disk boundaries. We briefly discuss potential applications of the 
equivalence between isothermal dusty gas and standard hydrodynamics
to provide thermodynamical interpretations of  other dusty
instabilities in protoplanetary disks.  
\end{abstract}

@@

\input{model}
\input{linear}
\input{result}
%\input{linear_adia}
%\input{numerical}
%\input{application} 
\input{summary}

### 2016-08-02 result.tex
adding fixZ results

We find that increasing the dust-to-gas ratio reduce VSI growth
rates. Notably `surface modes', which are typically fastest growing in
the dust-free case, are surpressed in dusty disks for $\epsilon_0\geq
0.1$. The body modes' growth rates remain $\sim O(h_\mathrm{g}\OmK)$ but their
oscillation frequency increases with $\epsilon_0$. The lowest
frequency `fundamental' body mode is not the fastest growing, but they
are energetically dominant because the entire disk column is perturbed
\citep[cf. surface modes which only disturb the disk boundaries,][]{umurhan16c}.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_eigenvals_kx30Hd1} 
  \caption{Unstable modes in a locally isothermal, perfectly coupled
    dusty disk with fiducial parameters
    $(p,q,h_\mathrm{g}, \Hd/\Hg )=(-1.5,-1,0.05, 0.99)$. The real
    frequency $\omega$ and growth rates $s$ are shown for a range of
    midplane dust-to-gas ratios $\epsilon_0=\rhod/\rhog$. 
    \label{vsi_dust_loading}
    }
\end{figure}

In Fig. \ref{vsi_dust_loading2d}
we compare the fundamental mode between the nearly 
dust-free case $\epsilon_0=10^{-3}$ an a dusty disk with
$\epsilon_0=1$. Dust-loading preferentially
stabilizes the disk atmosphere against the VSI, restricting
meridional motions to $|z|\lesssim 2\Hg$, where most of the disk mass
reside. Notice also the perturbed dust-to-gas ratio,
$\delta\epsilon$, shifts from being maximized at the disk boundaries
to the midplane.

@@

\subsection{Fixed metalicity}

>>>>

We find that increasing the dust-to-gas ratio reduce VSI growth
rates. Notably `surface modes', which are typically fastest growing in
the dust-free case, are surpressed in dusty disks for $\epsilon_0\geq
0.1$. The body modes' growth rates remain $\sim O(h_\mathrm{g}\OmK)$
but their oscillation frequency increases with
$\epsilon_0$. The total number of modes do not change
significantly. This is in contrast with the effect of increasing
cooling times, where \citetalias{lin15} find fewer unstable modes.
%until the system is eventually completely stable.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_eigenvals_kx30Hd1} 
  \caption{Unstable modes in a locally isothermal, perfectly coupled
    dusty disk with fiducial parameters
    $(p,q,h_\mathrm{g}, \Hd/\Hg )=(-1.5,-1,0.05, 0.99)$. The real
    frequency $\omega$ and growth rates $s$ are shown for a range of
    midplane dust-to-gas ratios $\epsilon_0=\rho_\mathrm{g0}/\rho_\mathrm{d0}$. 
    \label{vsi_dust_loading}
    }
\end{figure}

The lowest frequency `fundamental' body mode is not the fastest
growing, but they are energetically dominant because the entire disk
column is perturbed \citep[cf. surface modes which only disturb the
  disk boundaries,][]{umurhan16c}. In Fig. \ref{vsi_dust_loading2d}
we compare the fundamental mode between the nearly 
dust-free case $\epsilon_0=10^{-3}$ an a dusty disk with
$\epsilon_0=1$. Dust-loading preferentially
stabilizes the disk atmosphere against the VSI, restricting
meridional motions to $|z|\lesssim 2\Hg$, where most of the disk mass
reside. Notice also the perturbed dust-to-gas ratio,
$\delta\epsilon$, shifts from being maximized at the disk boundaries
to the midplane.

@@

\subsection{Fixed metalicity} 
Here we vary the dust layer thickness $\Hd$ but hold 
\begin{align}\label{metalicity_def}
Z \equiv \epsilon_0 \Htilde = \text{constant,}
\end{align}
so that $\Sigma_\mathrm{d}/\Sigma_\mathrm{g}$ is
fixed. Eq. \ref{metalicity_def} then gives the required midplane
dust-to-gas ratio.

%zmax = 2 because we have very thin dust layer

We set $Z=0.03$ and consider two disks with $\Hd=0.1\Hg$ and
$\Hd=0.99\Hg$. Fig. \ref{compare_vshear_fixZ} compares the

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_vshear_fixZ} 
  \caption{Vertical shear rate in a locally isothermal, dusty disk
    with fixed dust-to-gas mass and dust thickness $\Hd=0.1\Hg$
    (black) and $\Hd=0.99\Hg$ (blue). 
    \label{compare_vshear_fixZ}
    }
\end{figure}

\begin{figure}
  \includegraphics[scale=0.33, clip=true, trim=0.5cm 0cm 3cm 0cm]{figures/result2d_Hd1.ps}\includegraphics[scale=0.33, clip=true, trim=2cm 0cm 0cm 0cm]{figures/result2d_Hd0d1.ps} 
  \caption{
%dusty VSI mode in real space for midplane dust-to-gas
%    ratio $\epsilon_0=10^{-3}$ (top) and $\epsilon_0=1$
%    (bottom). The color scale shows the perturbation to the
%    dust-to-gas ratio, $\delta\epsilon$; and the arrows show
%    $\sqrt{\rho}\left(\dd v_x, \dd v_z\right)$. 
    stuff
    \label{result2d_fixZ}
    }
\end{figure}
